% Part: exercises
% Chapter: many-valued logic
% Section: semantics

\documentclass[../../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{exe}{mvl}{tsm}

\olsection{Three valued logic: semantics}

\begin{prob}[Evaluations]
Suppose that $\pValue{v}(\Obj p)=\True$, $\pValue{v}(\Obj q)=\Undef$, $\pValue{v}(\Obj r)=\False$.
Give the value of the following in three-valued logics:
\begin{enumerate}
\item (in $\LogKs$) $\Obj{p \lor q}$
\item (in $\LogKw$) $\Obj{p \lor q}$
\item (in $\LogKs$) $\Obj{p\land(q\lor r)}$
\item (in $\LogKs$) $\Obj{\lnot(q\lor r)\land p}$
\item (in $\LogKw$) $\Obj{\lnot(q\lor r)\land p}$
\item (in $\LogKs$) $\Obj{q\land\lnot q}$
\item (in $\LogKs$) $\Obj{q\lor\lnot q}$
\item (in $\LogLuk[3]$) $\Obj{q\rightarrow(p\lor r)}$
\item (in $\LogKs$) $\Obj{(p\land q)\rightarrow(p\lor r)}$ 
\item (in $\LogLuk[3]$) $\Obj{q\rightarrow(p\land r)}$
\item (in $\LogKs$) $\Obj{(p\land q)\rightarrow(q\lor r)}$ 
\item (in $\LogLuk[3]$) $\Obj{(p\land q)\rightarrow(q\lor r)}$
\end{enumerate}
\begin{ans}
\begin{enumerate}
	\item $\True$
	\item $\Undef$
	\item $\Undef$
	\item $\Undef$
	\item $\Undef$
	\item $\Undef$
	\item $\Undef$
	\item $\True$
	\item $\True$
	\item $\Undef$
	\item $\Undef$
	\item $\True$
\end{enumerate}
\end{ans}
\end{prob}

\begin{prob}[Equivalences between connectives]
Answer the following. 
\begin{enumerate}
\item Can you express $\land$ using only $\lor$ and $\lnot$ in the
	logics $\LogKw$, $\LogKs$, $\LogLuk[3]$? If yes, how? If no, explain 
	briefly why.
\item Can you express $\lor$ using only $\land$ and $\lnot$ in the logics 
	$\LogKw$, $\LogKs$, $\LogLuk[3]$? If yes, how? If no, explain briefly 
	why.
\item Can you express $\lif$ using only $\land$ and $\lnot$ in 
	$\LogKs$? If yes, how? If no, explain briefly why.
\item Can you express $\lif$ using only $\land$ and $\lnot$ in
	$\LogLuk[3]$? If yes, how? If no, explain briefly why.
\end{enumerate}
\begin{ans}
\begin{enumerate}
\item Yes. $!A \land !B$ is equivalent to $\lnot(\lnot !A \lor \lnot !B)$
	 in the logics $\LogKw$, $\LogKs$, $\LogLuk[3]$. This can be seen from
	 their respective truth-tables (draw the two truth-tables for the logic
	 $\LogKw$, the two for $\LogKs$, and point out that the semantics
	 for $\land$, $\lnot$, $\lor$ is the same in $\LogLuk[3]$.)
\item Yes. $!A \lor !B$ is equivalent to $\lnot(\lnot !A \land \lnot !B)$ in 
	the logics $\LogKw$, $\LogKs$, $\LogLuk[3]$. 
\item Yes. $!A \lif !B$ is equivalent to $\lnot (!A \lnot !B)$ in $\LogKs$. 
\item No. Consider a valuation $\pAssign{v}$ with $\pAssign{v}(\alpha)=\Undef$ 
	for any atomic !!{formula} $\alpha$. It is easy to see from the matrices for $\land$ 
	and $\lnot$ that every !!{formula} composed only of atomic !!{formula}s combined with $\land$
	and $\lnot$ will be $\Undef$ too. But $\Obj p \lif \Obj p$ is $\True$ on that valuation
	in $\LogLuk[3]$. So no !!{formula} containing only $\land$ and $\lnot$ can 
	express $\lif$ in $\LogLuk[3]$. 
\end{enumerate}
\end{ans}
\end{prob}

\begin{prob}[Expressive power (advanced, optional)]
Answer the following.
\begin{enumerate}
	\item Is there, in any of $\LogKw$, $\LogKs$, $\LogLuk[3]$, a !!{formula} that is $\True$ if $\Obj p$ is $\True$, and $\False$ otherwise? (Note that $\Obj p$ itself is $\True$ if $\Obj p$ is $\True$, but not $\False$ otherwise, since it is not $\False$ when $\Obj p$ is $\Undef$).
	\item If yes, say which !!{formula} in which logic.
	\item If no, show that it is so. 
\end{enumerate}
\begin{ans}
\begin{enumerate}
	\item Not in $\LogKw$, $\LogKs$, but yes $\LogLuk[3]$.
	\item In $\LogLuk[3]$, $\Obj{\lnot (p \lif \lnot p)}$.
	\item In $\LogKw$, $\LogKs$. Consider the !!{valuation} with $\pAssign{v}(\alpha)=\Undef$ for any !!{propositional variable} $\alpha$ (so $\pAssign{v}(\Obj p)=\Undef$). In that !!{valuation}, $\pValue{v}(!A)=\Undef$ for any formula $!A$. If there was !!a{formula} that is $\False$ whenever $\Obj p$ is $\Undef$, that !!{formula} 
	would be $\False$ in $\pAssign{v}$, so there is no such !!{formula}. 
	In $\LogLuk[3]$, consider the truth-table for $\Obj{\lnot (p \lif \lnot p)}$:
	\begin{center}
	\begin{tabular}{c|ccc}
	\hline
	$\Obj{p}$ & $\Obj{\lnot p}$ & $\Obj{p \lif \lnot p}$ & $\Obj{\lnot (p \lif \lnot p)}$\\
	\hline
	$\True$ & $\False$ & $\False$ & $\True$ \\
	$\Undef$ & $\Undef$ & $\True$ & $\False$ \\
	$\False$ & $\True$ & $\True$ & $\False$ \\
	\hline
	\end{tabular}
	\end{center} 
\end{enumerate}
\end{ans}
\end{prob}

\begin{prob}[Validity]
Say which of the following hold in $\LogKs$ and in $\LogLuk[3]$. For each, give a truth-table or an informal proof. If the table or proof is the same for both logics, only give one. 
\begin{enumerate}
	\item $\Obj{\Entails p \lor\lnot p}$
	\item $\Obj{\Entails \lnot(p \land\lnot p)}$
    \item $\Obj{p, p \lif q \Entails q}$
    \item $\Obj{\lnot\lnot p \Entails p}$
    \item $\Obj{\Entails p \rightarrow p}$
    \item $\Obj{p \land q \Entails p}$
    \item $\Obj{p \Entails p \lor q}$
    \item $\Obj{p, \lnot p \Entails q}$
    \item $\Obj{\Entails (p \land \lnot p) \rightarrow q}$
\end{enumerate}
\begin{ans}
	!!^{formula}s that do not have $\lif$ (nor $\liff$) have the same value
	 in $\LogKs$ and $\LogLuk[3]$ on any !!{valuation}. So when the 
	 argument does not involve $\lif$ it is enough to consider
	only one of these logics; the argument is valid in one just if it 
	is valid on the other.
\begin{enumerate}
\item Incorrect in $\LogKs$ and $\LogLuk[3]$. Consider $\pAssign{v}$
	with $\pAssign{v}(\Obj p)=\Undef$ in $\LogKs$, we have
	$\pValue{v}(\Obj{\lnot p})=\Undef$ and thus $\pValue{v}(\Obj{p \lor 
	\lnot p})=\Undef$.
\item Incorrect in  $\LogKs$ and $\LogLuk[3]$. Consider $\pAssign{v}$
	with $\pAssign{v}(\Obj p)=\Undef$ in $\LogKs$, we have
	$\pValue{v}(\Obj{\lnot p})=\Undef$, hence $\pValue{v}(\Obj{p
	\land\lnot p})=\Undef$, hence $\pValue{v}(\Obj{\lnot(p \land\lnot
	p)})=\Undef$. item Correct in $\LogKs$. Suppose
	$\pValue{v}(\Obj{p})=\True$, $\pValue{v}(\Obj{p\lif q})=\True$ on a
	$\LogKs$ !!{valuation}. Then $\pValue{v}(\Obj{p\lif q})=\True$ only if
	$\pValue{v}(\Obj{p})=\False$ or $\pValue{v}(\Obj{q})=\True$; since
	$\pValue{v}(\Obj{p})\neq \False$, $\pValue{v}(\Obj{q})=\True$. So
	there is no !!{valuation} under which the premises are $\True$ but the
	conclusion $\Undef$ or $\False$. 
\item Correct in $\LogKs$ and
	$\LogLuk[3]$, as shown by the table below (adequate for both
	logics).
	\begin{center}
	\begin{tabular}{|c|cc|}
	\hline
	$\Obj p$ 	&	$\Obj{\lnot p}$		&	$\Obj{\lnot \lnot p}$\\
	\hline
	$\True$		&	$\False$			&	$\True$\\
	$\Undef$	&	$\Undef$			&	$\True$\\
	$\False$	&	$\True$				&	$\True$\\
	\hline
	\end{tabular}
	\end{center}
\item Incorrect in $\LogKs$: in $\pAssign{v}$ with $\pAssign{v}(\Obj
	p)=\Undef$ we have $\pValue{v}(\Obj {p \lif p})=\Undef$ in the $\LogKs$
	semantics.\\
	Correct in $\LogLuk[3]$: $p \lif p$ is $\True$ in the $\LogLuk[3]$
	semantics if $p$ is $\True$, $\Undef$ or $\False$.
\item Correct in $\LogKs$ and $\LogLuk[3]$, as shown by the 
	table below (adequate for both logics).
	\begin{center}
	\begin{tabular}{|cc|c|}
	\hline
	$\Obj p$	&	$\Obj q$	&	$\Obj{p \land q}$	\\
	$\True$		&	$\True$		& 		$\True$ 		\\
	$\True$		&	$\Undef$ 	& 		$\Undef$		\\
	$\True$		&	$\False$ 	& 		$\False$ \\
	$\Undef$	&	$\True$		& 		$\Undef$ 		\\
	$\Undef$	&	$\Undef$ 	& 		$\Undef$		\\
	$\Undef$	&	$\False$ 	& 		$\False$ \\
	$\False$	&	$\True$		& 		$\False$ 		\\
	$\False$	&	$\Undef$ 	& 		$\False$		\\
	$\False$	&	$\False$ 	& 		$\False$ \\
	\hline
	\end{tabular}
	\end{center}
	There is no line at which the premise is true and the conclusion
	not true. (There is only one line at which the premise is true, 
	the first line, and the conclusion is true at that line too.)
\item Correct in $\LogKs$ and $\LogLuk[3]$. Consider any $\pAssign{v}$ 
	such that $\pValue{v}(\Obj p)=\True$. Then $\pValue{v}(\Obj{p \lor q})=\True$ 
	on any many-valued logic. So there is no !!{valuation} that makes
	the premise true and the conclusion not true. 
\item Correct in $\LogKs$ and $\LogLuk[3]$. There is no !!{valuation}
	$\pAssign{v}$ under which both $\Obj p$ and $\Obj \lnot p$ are 
	$\True$. A fortiori no !!{valuation} under which the premises are 
	$\True$ and the conclusion $\Undef$ or $\False$.
\item Incorrect in $\LogKs$, correct in $\LogLuk[3]$.\\
	For $\LogKs$, consider $\pAssign{v}$ with 
	$\pAssign{v}(\Obj p)=\Undef$ and $\pAssign{v}(\Obj p)=\False$. 
	We have $\pValue{v}(\Obj{p \land \lnot p})=\Undef$ and 
	$\pValue{v}(q)=\False$, so 
	$\pValue{v}(\Obj{(p \land \lnot p \lif q)})=\Undef$. Hence there
	are !!{valuation}s under which the !!{formula} is not true in $\LogKs$.\\
	In $\LogLuk[3]$, the truth-table below shows that the formula
	is $\True$ under any !!{valuation}.
	\begin{center}
	\begin{tabular}{|cc|ccc|}
	\hline
	$\Obj p$	&	$\Obj q$	&	$\Obj{\lnot p}$ &	$\Obj{p \land \lnot p}$	&	$\Obj{(p \land \lnot p) \lif q}$ \\
	$\True$		&	$\True$		& 	$\False$		&		$\False$ 			&	$\True$ \\
	$\True$		&	$\Undef$ 	&	$\False$		& 		$\False$ 			&	$\True$ \\
	$\True$		&	$\False$ 	&	$\False$		& 		$\False$ 			&	$\True$ \\
	$\Undef$	&	$\True$		&	$\Undef$		& 		$\Undef$  			&	$\True$ \\
	$\Undef$	&	$\Undef$ 	&	$\Undef$		& 		$\Undef$ 			&	$\True$ \\
	$\Undef$	&	$\False$ 	&	$\Undef$		& 		$\Undef$ 			&	$\True$ \\
	$\False$	&	$\True$		&	$\True$			& 		$\False$ 			&	$\True$ \\
	$\False$	&	$\Undef$ 	&	$\True$			& 		$\False$ 			&	$\True$ \\
	$\False$	&	$\False$ 	&	$\True$			& 		$\False$ 			&	$\True$ \\
	\hline
	\end{tabular}
	\end{center}
	\emph{Comment.} At lines 4-6 you would get the last !!{formula} $\Undef$ in the semantics
	for $\LogKs$ and $\LogLP$.
\end{enumerate}
\end{ans}
\end{prob}

\begin{prob}[Sub-classicality]
Pick one of $\LogKw$, $\LogKs$, $\LogLuk[3]$ as $\Log{X}$. 
Show that the following holds: 
\begin{itemize}
	\item If $\Gamma \Entails[\Log{X}] !A$ then  $\Gamma \Entails[\LogCL] !A$.
\end{itemize}
\emph{Tip}. Prove the converse. Suppose that $\Gamma \Entails/[\LogCL] !A$, 
explain why $\Gamma \Entails/[\Log{X}] !A$, with your chosen $\Log{X}$.
\begin{ans}
\emph{Comment: in the following you can replace $\Log{X}$ with any of
$\LogKw$, $\LogKs$, $\LogLuk[3]$: the answer is the same.}

Suppose $\Gamma \Entails/[\LogCL] !A$. Then there is a \emph{classical} 
!!{valuation} $\pAssign{v}$ under which all !!{formula}s of $\Gamma$ are 
$\True$ but $!A$ false. Now a classical !!{valuation} is also a !!{valuation}
for the logics $\Log{X}$ with all the atoms assigned $\True$ or $\False$. 
Moreover, it is easy to see from the semantics of $\Log{X}$ that when a
!!{valuation} that only assigns $\True$ or $\False$ to all atoms, $\Log{X}$
agrees with $\LogCL$ on all !!{formula}s (that is, the value $\Log{X}$ assigns
to complex !!{formula}s is the same as the one assigned by $\LogCL$
on the basis of the same !!{valuation}). So there is a !!{valuation}
in $\Log{X}$ that makes the premises $\True$ but the conclusion not $\True$.
Hence $\Gamma \Entails/[\Log{X}] !A$. 

\emph{OPTIONAL. If you are enthusiastic you can add a proof of the second
step, by induction on the length of !!{formula}s. This goes as follows, but
is not required for a correct answer.} Let $\pAssign{v}$ be a $\LogCL$ 
!!{valuation} and $\pAssign{u}$ be a $\Log{X}$ !!{valuation} with
$\pAssign{u}(\alpha)=\pAssign{v}(\alpha)$ for any !!{propositional variable}
$\alpha$. Since $\pAssign{v}$ is a classical !!{valuation} we have 
$\pAssign{v}(\alpha)=\True$ or $\False$ for any !!{propositional variable} $\alpha$.
We prove by induction on the length of !!{formula}s that $\pValue{u}(!A)
=\pValue{v}(!A)$ for all $!A$.

Base case. $!A$ is an atom. Trivial ($\pValue{u}(!A) = \pAssign{u}(!A)
=\pAssign{v}(!A) = \pValue{v}(!A)$).

Induction hypothesis: suppose that for any !!{formula} $!B$ shorter than $!A$,
$\pValue{u}(!B) = \pValue{v}(!B) = \True$ or $\False$.

\begin{enumerate}
\item \indcase{!A}{\alpha}{As in the base case.}
\item \indcase{!A}{\lnot !B}{By the induction hypothesis,
$\pValue{u}(!B) = \pValue{v}(!B)$ and $\pValue{v}(!B) = \True$ or $\False$.
It is easy to see from the matrix of $\lnot$ in $\Log{X}$ that we will have
$\pValue{u}(\lnot !B) = \pValue{v}(\lnot !B)$, that is, $\pValue{u}(!A) 
= \pValue{v}(!A)$.}
\item \indcase{!A}{!B \lif !C}{By the induction hypothesis,
$\pValue{u}(!B) = \pValue{v}(!B)$ and $\pValue{v}(!B) = \True$ or $\False$.
It is easy to see from the matrix of $\lif$ in $\Log{X}$ that we will have
$\pValue{u}(!B \lif !C) = \pValue{v}(!B \lif !C)$, that is, $\pValue{u}(!A) 
= \pValue{v}(!A)$.}
\item The cases $!A\ident !B\land!C$ and $!A\ident !B\lor !C$ are similar.
\end{enumerate}
This completes the induction. 
\end{ans}
\end{prob}

\begin{prob}[No valid !!{formula}s in Kleene logics]
Explain why there are no valid !!{formula}s in $\LogKs$ and $\LogKw$.
\begin{ans}
Consider !!a{valuation} that assigns $\Undef$ to every !!{propositional
variable} $\alpha$. In the semantics of $\LogKs$ and $\LogKw$, 
if $!A$ and $!B$ are $\Undef$ (in some !!{valuation}) then 
$\lnot !A$, $!A \land !A$, $!A \lor !B$
are $\Undef$ too. From this it can be seen that every complex !!{formula}
will have the value $\Undef$ in the !!{valuation} $v$ too. Hence no 
!!{formula} is true in every valuation in $\LogKs$, nor in $\LogKw$. 

\emph{Comment. You could prove this rigorously by induction on the
length of !!{formula}s. This is shown below, but the above is sufficient
for a correct answer}. Let $\pAssign{v}$ be such that $\pAssign{v}(\alpha)
= \Undef$ for every !!{propositional variable} $\alpha$. We show by
induction on the length of formulas that in $\LogKs$ and $\LogKw$,
 $\pValue{v}(!A)=\Undef$ forany !!{formula} $!A$.

Base case. $!A$ is a !!{propositional variable}. $\pValue{v}(!A)=
\pAssign{v}(!A)=\Undef$.

Induction hypothesis: suppose that for any !!{formula} $!B$ shorter than 
$!A$ we have $\pValue{v}(!B)=\Undef$.

\begin{enumerate}
	\item \indcase{!A}{\alpha}{As in the base case.}
	\item \indcase{!A}{\lnot !B}{Since $\pValue{v}(!B)=\Undef$ (induction
	hypothesis), $\pValue{v}(\lnot !B)=\Undef$ in $\LogKs$ and $\LogKw$.}
	\item \indcase{!A}{!B \lif
    !C}{Since $\pValue{v}(!B)=\Undef$ and $\pValue{v}(!C)=\Undef$ (induction
	hypothesis), $\pValue{v}(!B \lif !C)=\Undef$ in $\LogKs$ and $\LogKw$.}
	\item The cases for $!A\ident !B \land !C$ and $!A\ident !B \lor !C$
	are similar.
\end{enumerate}
This completes the induction. 
\end{ans}
\end{prob}

\begin{prob}[Non-contradiction and non-tautologies in $\LogLuk$]
Considering $\LogLuk[3]$:
\begin{enumerate}
\item Give !!a{formula} that is valid.
\item Give a !!{formula} that is not valid, but cannot be false under
any  !!{structure}.
\item Give !!a{formula} that cannot be true under any !!{structure}, but 
that is not a contradiction (i.e., its negation is not valid).
\end{enumerate}
\begin{ans}
\begin{enumerate}
\item $\Obj{p \lif p}$\\
	\emph{Comment}. There are many options, e.g.  $\Obj{(p \land q) \lif
	q}$, $\Obj{p \lif (q \lor p)}$,  $\Obj{(p \land \lnot p) \lif q}$. They
	must all involve the conditional, since a formula with only $\lnot$,
	$\lor$, $\land$ will be $\Undef$ on the !!{valuation} that assigns
	$\Undef$ to all !!{propositional variable}s. But in need not have
	$\lif$ as the !!{main operator}: $\Obj{(p\lif p)\lor \lnot(p\lif p)}$
	is valid too.  
\item $\Obj{p \lor \lnot p}$
\item $\Obj{p \land \lnot p}$\\
	\emph{Comment}. In $\LogLuk[3]$ (as in $\LogKw$, $\LogKs$
	$\LogLP$), $\pValue{v}(\Obj{p \land \lnot p}) = \False$ if 
	$\pValue{v}(\Obj{p})=\True$ or $\False$, and $\Undef$ if 
	$\pValue{v}(\Obj{p}=\Undef)$. Hence the !!{formula} cannot be
	true, but its negation is not valid in $\LogLuk$ because it is
	$\Undef$ under !!{valuation}s such that $\pValue{v}(\Obj{p \land 
	\lnot p}) = \Undef$.
\end{enumerate}
\end{ans}
\end{prob}

\begin{prob}[Conditional and consequence]
Consider the following claim: 

$!A, !B \Entails !C$ if, and only if, $!A \Entails !B \lif !C$.
\begin{enumerate}
	\item Does this hold in $\LogKs$? If yes, show why, if not, say which
	 direction (if, only if, both) fails and give the relevant counterexamples.
	 \item Same question for $\LogLuk[3]$.
\end{enumerate}
\begin{ans}
\begin{enumerate}
\item $\LogKs$. The left to right (if) direction holds. Suppose $!A \Entails !B \lif
	!C$.  Let $\pAssign{v}$ be any !!{valuation} such that
	$\pValue{v}(!A)=\True$ and $\pValue{v}(!B)=\True$. Since $!A \Entails
	!B \lif !C$, any !!{valuation} in which $!A$ is $\True$ is such that
	$!B \lif !C$ is $\True$, so $\pValue{v}(!B  \lif !C)=\True$. By the
	$\LogKs$ semantics, $\pValue{v}(!B)=\True$ and  $\pValue{v}(!B \lif
	!C)=\True$ only if $\pValue{v}(!C)=\True$. So there is no
	!!{valuation} $\pAssign{v}$ such that $\pValue{v}(!A)=\True$ and
	$\pValue{v}(!B)=\True$ and $\pValue{v}(!C)$ not $\True$: $!A, !B
	\Entails !C$.\\ 
	The right to left (only if) direction fails.
	Counterexample: $\Obj{q},  \Obj{p} \Entails \Obj{p}$ (trivially, no
	!!{valuation} that makes $\Obj{p,q}$ $\True$ will make $\Obj{p}$ non
	true), but $\Obj{q} \Entails/ \Obj{p \lif p}$ (consider  consider
	$\pAssign{v}$ with $\pAssign{q}=\True$ and $\pAssign{p}=\Undef$).\\
	\emph{Comment: See the footnote for the reasoning that leads me to
	the right to left counterexample.}\footnote{First I note $!A, 
	!B \Entails[\LogKs]
	!C$ turns on what happens on !!{valuation}s where \emph{both} $!A$
	and $!B$ are $\True$, while $!A \Entails[\LogKs] !B \lif
	!C$ turns on \emph{all} !!{valuation}s where $!A$ is true. So we can
	infer that if the former doesn't entail the latter, it's going to 
	have to do with the !!{valuation}s where ($!A$ is true but) $!B$ is $\Undef$ 
	or $\False$ rather than $\True$. (I can narrow it down to those
	where $!A$ is $\Undef$ by noting that $!B \lif !C$ will be $\True$ 
	whenver $!B$ is $\False$, so the !!{valuation}s with $!B$ $\False$ 
	cannot make the right-side entailment fail.) 

	Second, I look 
	at what would have to be the case for that direction
	 to \emph{fail}. It would have to be that: every !!{valuation} with 
	 $!A$ and $!B$ $\True$
	has $!C$ $\True$ too, but some !!{valuation} $\pAssign{v}$ with $!A$ 
	$\True$ makes $!B \lif !C$ $\Undef$ or $\False$. Now we know $\pAssign{v}$
	couldn't make $!B$ $\True$ (since every !!{valuation} that makes both $!A$
	and $!B$ $\True$ makes $!C$ $\True$ and that would make $!B \lif !C$) $\True$).
	And we know that $\pAssign{v}$ couldn't make $!B$ $\False$, because 
	that would make $!B \lif !C$ $\True$. So it would have to be a 
	!!{valuation} that makes $!B$ $\Undef$, and $!B \lif !C$ $\Undef$
	of $\False$. From the table of $\lif$ in $\LogKs$ we can see that
	with $!B$ $\Undef$ so our only options is to get $!B \lif !C$ $\Undef$
	with $!C$ either $\Undef$ or $\False$. 

	So, third, I look for three !!{formula}s $!A$, $!B$,
	$!C$ such that $!B \lif !C$ can be $\Undef$ or $\False$ when $!B$ is
	$\Undef$ (and $!A$ $\True$). But my !!{formula}s should also be such
	that $!C$ is $\True$ when both $!A$ and $!B$ are. A simple solution is
	to let $!B \ident !C \ident \Obj{p}$, since in $\LogKs$ $\Obj{p \lif
	p}$ is $\Undef$ when $\Obj{p}$ is.}
\item $\LogLuk[3]$.  The left to right (if) direction holds. The proof steps the
	same as in $\LogKs$. In particular, note that in the
	$\LogLuk[3]$ semantics too, $\pValue{v}(!B)=\True$ and  $\pValue{v}(!B \lif
	!C)=\True$ only if $\pValue{v}(!C)=\True$.\\
	The right to left direction fails. Counterexample: $\Obj{r}, \Obj{p  \land (p
	\lif q)} \Entails[\LogLuk[3]] q$ but $\Obj{r} \Entails/[\LogLuk[3]] \Obj{(p
	\land (p \lif q))\lif q}$. For the first, let $\pAssign{v}$ by any 
	!!{valuation} such that ($\pAssign{v}(\Obj{r})=\True$ and $\pValue{v}(\Obj{p
	\land (p \lif q)})=\True$. It is easy to see from the semantics that
	$\pValue{v}(\Obj{p})=\True$. For the second, consider $\pAssign{v}$ such that
	($\pAssign{v}(\Obj{r})=\True$ and) $\pAssign{v}(\Obj{p})=\Undef$ and 
	$\pAssign{v}(\Obj{q})=\False$. We have $\pValue{v}(\Obj{p \lif q})=\Undef$,
	$\pValue{v}(\Obj{p\land (p \lif q)})=\Undef$, and since $\pValue{v}(\Obj{q})=\False$, $\pValue{v}(\Obj{p\land (p \lif q)\lif q})=\Undef$.\footnote{Reasoning 
	that leads to the counterexample. I suspect that this direction may fail 
	for the reason as before: what happens when $!B$ is $\Undef$ rather than $\False$.
	So I ask myself what has to be the case for this direction to fail. It has to be that there
	are three !!{formula}s $!A,!B,!C$ such that (a) $!C$ is $\True$ on
	any !!{valuation} on which \emph{both} $!A, !B$ are $\True$ but 
	(b) $!B \lif !C$ is $\Undef$ or $\False$ on some valuation $\pAssign{v}$ 
	on which $!A$ is $\True$. We can rule out that $\pValue{v}(!B)=\True$
	(for then $\pValue{v}(!C)=\True$ by (a) and hence $\pValue{v}(!B 
	\lif !C)=\True$ too)
	and $\pValue{v}(!B)=\False$ (since then $\pValue{v}(!B 
	\lif !C)=\True$ too). But in $\LogLuk[3]$ if $\pValue{v}(!B)$ the only 
	way for $!B \lif !C$ to be other than $\True$ is to have 
	$\pValue{v}(!C)=\False$. (In $\LogKs$, having $!C$ $\Undef$ would be enough
	to make $!B\lif !C$ not true, but in $\LogLuk[3]$ with $!B$ $\Undef$ we 
	need $!C$ \False.)\\
	So I am looking for a pair of !!{formula}s $!B$, $!C$ such that
	in every valuation in which $!B$ (and $!A$) is $\True$, $!C$ is 
	$\True$, and for some valuation $\pAssign{v}$, $\pValue{v}(!B)=\Undef$
	and $\pValue{v}{!C}=\False$. (Or at least two formulas that do so on the 
	background of some formula $!A$.)\\
	Here some exploration and trial and error is needed. What put me on the
	track was the thought that in $\LogLuk[3]$ a $\True$ conditional $\Obj{p \lif q}$
	doesn't allow any `drop' in truth-value from $\Obj p$ to $\Obj q$, while a $\Undef$
	conditional allow from drops. So I looked at using $\Obj{p \lif q}$ for $!B$ and $\Obj q$ $!C$. But I needed the premises $!A, !B$ to entail $!C$ when true, so I needed to introduce $\Obj p$ as either $!A$ (but this doesn't work) or a conjunct in $!B$. Hence the idea of having $\Obj{p \land (p \lif q)}$ as my premise $!B$. When fully true $\True$, it makes both $\Obj{p}$ and $\Obj{p \land q}$ $\True$, so 
	no drop of truth-value from $\Obj p$ to $\Obj q$ and $\Obj{q}$ is $\True$. But when $\Obj{p \land (p \lif q)}$ is not $\True$ but $\Undef$, we can have $\Obj{p \lif q}$ itself $\Undef$ and allow a drop in truth-value from from $\Obj p$ to $\Obj q$. I thus check what happens with $\Obj p$ $\Undef$ and $\Obj q$ $\False$ and 
	find out that the couterexample works.}
\end{enumerate}
\end{ans}
\end{prob}

\end{document}