% Part: quantified-modal-logic
% Chapter: introduction
% Section: freges-problem

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{qml}{int}{frg}

\olsection{Frege's Problem}

We use modal operators to express not only necessity but also epistemic
notions, like knowledge, belief, and others. For instance, we may 
introduce a unary operator of propositions $K$ to talk about what a 
certain subject knows: $K!A$ states that `one' (that subject) `knows
that $!A$'. 

Frege's problem arises from a simple application of Leibniz's law. 
Informally, if $a$ is $b$ then whatever is true of $a$ is true of 
$b$ (Leibniz's law). If, in particular, it is true of $a$ that 
one knows it to be $a$, then it is true of $b$ that one knows it 
to $b$. So if one knows that $a$ is $a$ then one knows that $b$ is 
$a$. 

Formally, the proof only uses Leibniz's law and propositional logic:
\begin{prooftree}
    \AxiomC{$\Discharge{\eq[a][b]}{2}$}
        \AxiomC{$\Discharge{K\eq[a][a]}{1}$}
    \RightLabel{\Elim{\eq}}
    \BinaryInfC{$K\eq[a][b]$}
    \RightLabel{\Intro{\rightarrow}}
    \UnaryInfC{$K\eq[a][a]\rightarrow K\eq[a][b]$}
    \RightLabel{\Intro{\rightarrow}}
    \UnaryInfC{$\eq[a][b]\rightarrow(K\eq[a][a]\rightarrow K\eq[a][b])$}
\end{prooftree}

On the face of it, the result is paradoxical. Even though Jay-Z is 
Shaw Carter, many people who know that Jay-Z is Jay-Z don't know 
that Shawn Carter is Jay-Z. Or so it seems. Yet the !!{derivation} 
concludes otherwise: everyone who knows that Jay-Z is Jay-Z knows 
that Shawn Carter is Jay-Z.

Of course, we could go further. With the Reflexivity of identity and 
Necessitation we derive $K\eq[a][a]$ from no premises. We can then 
simplify our conclusion to $\eq[a][b]\rightarrow K\eq[a][b]$. (Namely,
repeat the proof of Necessity of Identity with $K$ instead of $\Box$.)
However, it's worth stressing that the extra step is inessential to 
the main problem. Perhaps $\ulcorner K\eq[a][a] \ulcorner$ is a 
logical truth and anyone knows that Jay-Z is Jay-Z. Or perhaps we 
should reject one of Necessitation or Reflexivity and deny that 
everyone knows that Jay-Z is Jay-Z. Perhaps the people who have 
never heard of him do not know that. No matter. All the problem needs
is first, that there are in fact some people who know that Jay-Z is 
Jay-Z. And surely there are. And second, that some of these people 
plausibly don't know that Jay-Z is Shawn Carter. These two things 
are enough to generate a problem. For they conflict with Leibniz's 
law.  

\end{document}
