% Part: quantified-modal-logic
% Chapter: introduction
% Section: what-is-qml

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{qml}{int}{wiqml}

\olsection{What is Quantified Modal Logic?}

Quantified modal logics combine first-order logic, usually with
 identity, with modal logics. \emph{First-order logic with identity}
 can express claims about particular things: how things are and how
 they are related (subject-predicate sentences), whether something is
 the same thing as something else (identity), and, to some extent, how
 many things there are (e.g., for any natural number $n$, the claims
 that there are at least $n$ things). Modal logics can express claims
 about whether something is necessarily so, possibly so but also
 claims about what is known to be so, knowned to be so, believed to be
 so, hoped to be so and so on. Many claims of philosophical interest
 combine notions of both lists, and can only be captured by combining
 the resources of both logics:

\begin{itemize}
\item Annie is necessarily human. $\Box\Atom{H}{a}$
\item Every human is necessarily human. $\lforall[x][\Atom{H}{x}\lif\Box\Atom{H}{x}]$
\item God necessarily exists. $\Box\lexists{x}{\eq[a][x]}$
\item There could have been nothing. $\Diamond\lnot\lexists[x][\eq[x][x]]$
\item Everything is necessarily physical. $\lforall[x][\Box\Atom{P}{x}]$
\item If $a$ is $b$ then necessarily $a$ is $b$. $\eq[a][b]\lif\Box(\eq[a][b])$
\item It is a priori that all humans are humans.
$A(\Atom{H}{a}\lif\Atom{H}{a})$ (with $A$ a $\Box$-like operator 
for `it is a priori that \dots')
\item If one knows that Phosphorus shines, and one doesn't knows that
Hesperus shines, then one doesn't know that Hesperus is Phosphorus.
$(K\Atom{P}{a}\land\lnot K\Atom{P}{b})\lif\lnot K\eq[a][b]$ (with $K$ 
a $\Box$-like operator for `one knows that \dots')
\end{itemize}

Accordingly, Quantified Modal Logics are central to several important
philosophical debates. 

This introduction introduces major philosophical issues arising from
Quantified Modal Logic.

\end{document}