% Part: formatives
% Chapter: one
% Section: propositional-modal-logic

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{for}{one}{pmod}

\newpage
\olsection{Propositional Modal Logic}

\begin{prob}
 Consider the model $\mModel{M}$ with:
 \begin{itemize}
    \item Worlds $W=\{@,u,v\}$, where $@$ is the actual world,
    \item Accessibility relation 
    $R=\{\tuple{@,u},\tuple{@,v},\tuple{u,@},\tuple{v,u},\tuple{v,v}\}$
    \item Valuation $V(\Obj{\pvar{P}},@)=V(\Obj{\pvar{Q}},u)=V(\Obj{\pvar{R}},v)=\True$, all else $\False$.
 \end{itemize}
 Say whether the formulas below are true in the model (i.e., true at
    $@$). Briefly justify your answers.
    \begin{enumerate}
    \item $\Diamond \Obj{\pvar{P}}$
    \item $\Box(\Obj{\pvar{R}}\rightarrow\Diamond \Obj{\pvar{Q}})$
    \item $\Box(\Diamond \Obj{\pvar{\Obj{\pvar{R}}}}\lor\Diamond \Obj{\pvar{P}})$
    \item $\Box\Box\Diamond \Obj{\pvar{Q}}$
    \end{enumerate}
    
\end{prob}

\begin{prob}
    Establish the following. Use countermodels for invalidities, 
    natural deduction !!{derivation}s for validities. 
    \begin{enumerate}
        \item $\Box\lnot\Obj{\pvar{A}}, 
        \Box(\Obj{\pvar{A}}\lor\Obj{\pvar{B}})
        \Entails[\Log{K}] \Box{\Obj{\pvar{B}}}$

        \item $\Diamond\lnot\Obj{\pvar{A}}, 
        \Box(\Obj{\pvar{A}}\lor\Obj{\pvar{B}})
        \Entails/[\Log{T}] \Box{\Obj{\pvar{B}}}$

        \item $\Entails[\Log{T}]\Diamond(
            (\Obj{\pvar{A}}\land\Obj{\pvar{B}})\lif\Obj{\pvar{A}}
        )$

        \item $\Diamond\Diamond \Obj{\pvar{A}},
        \Box(\Obj{\pvar{B}}\lif\lnot\Obj{\pvar{A}})
        \Entails[\Log{S4}] \lnot\Box\Obj{\pvar{B}}$
    
    \end{enumerate}

\end{prob}

\begin{prob}
    We say that relation $R$ has \emph{secondary reflexivity} iff 
    whenever something has that relation to something, the latter has
    that relation to itself: if $Rww'$ then $Rw'w'$. A modal !!{structure}
    is secondary reflexive iff its accessibility relation is secondary 
    reflexive.

    \begin{enumerate}
    \item Can a model can be secondary reflexive without 
    being serial? Briefly justify your answer.
    \item Show that any transitive and symmetric model is secondary 
    reflexive, but not the other way round.
    \item Can you give a characteristic schema for secondary reflexive 
    frames? Show that the schema is valid for any secondary reflexive frames.
    Optional: show that it is not valid on frames that are not secondary 
    reflexive.
    \end{enumerate}
    
\end{prob}

\end{document}