% Part: exercise-booklet
% Chapter: propositional-modal-logic
% Section: semantics-DTB45

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{exb}{pml}{smd}

\olsection{Semantics in systems beyond $\Log{K}$}

\begin{prob}
    \citep[exercise 3.1 items 1-7]{MacFarlane-2020-PhilosophicalLogicContemporary}
    Find pointed !!{structure}s of the following kinds. 
    \begin{enumerate}
        \item Find a $\Log{K}$ !!{structure} in which
        $\Box\Obj{\pvar{A}}\lif\Diamond\Obj{\pvar{A}}$ is false.
        \item Find a $\Log{T}$ !!{structure} in which
        $\Box\Obj{\pvar{A}}\lif\Box\Box\Obj{\pvar{A}}$ is false.
        \item Find a $\Log{D}$ !!{structure} that is not a $\Log{T}$
        !!{structure}.
        \item Find a $\Log{S4}$ !!{structure} in which 
        $\Diamond\Obj{\pvar{A}}\lif\Box\Diamond\Obj{\pvar{A}}$ is false.
        \item Find a $\Log{S4}$ !!{structure} in which 
        $\Diamond\Box\Obj{\pvar{A}}\lif\pvar{A}$ is false.
        \item Find a $\Log{B}$ !!{structure} in which 
        $\Diamond\Obj{\pvar{A}}\lif\Box\Diamond\Obj{\pvar{A}}$ is false.
        \item Find a $\Log{S5}$ !!{structure} in which
        $\Obj{\pvar{A}}\lif\Box\Obj{\pvar{A}}$ is false.
    \end{enumerate}

    \begin{ans} In each case there are many (infinitely many) correct
        answers; I'll give the simplest ones below. In each case 
        I'll also explain which feature your !!{structure} must have to be 
        a correct answer. For the first few answers I'll also illustrate
        how you give a brief justification (if asked to do so): how you
        explain that the !!{structure}s are of the right type and 
        why the !!{formula} is false in them. 
    \begin{enumerate}
        \item A $\Log{K}$ !!{structure} in which
        $\Box\Obj{\pvar{A}}\lif\Diamond\Obj{\pvar{A}}$ is false.

        $\mModel{M}$ with $W=\{@\}$, $R=\emptyset$, $V$ any valuation.
        This is a $\Log{K}$ !!{structure} since there are no constraints on
        such !!{structure}s. At $@$ no world is accessible, so at $@$ any
        $\Box!A$ !!{formula} is true and any $\Diamond!A$ !!{formula}
        is false.

        Feature for correct answer: your !!{structure} needs a world
        that does not have access to any world.
        
        \item A $\Log{T}$ !!{structure} in which
        $\Box\Obj{\pvar{A}}\lif\Box\Box\Obj{\pvar{A}}$ is false.

        $\mModel{M}$ with $W=\{@,u,v\}$,
        $R=\{\tuple{@,@},\tuple{@,u},\tuple{u,u},\tuple{u,v},
        \tuple{v,v}\}$, $V(\Obj\pvar{A})=\{@,u\}$. That is a $\Log{T}$
        !!{structure} is $R$ is reflexive. Since $Ruv$ and
        $\mSat/{M}{\Obj\pvar{P}}[v]$,
        $\mSat/{M}{\Box\Obj\pvar{P}}[u]$. Since $R@u$ and
        $\mSat/{M}{\Box\Box\Obj\pvar{P}}[@]$. However, since $@,u$ are
        the only worlds accessible from $@$ and
        $\mSat{M}{\Obj\pvar{P}}[@]$ and $\mSat{M}{\Obj\pvar{P}}[u]$,
        $\mSat{M}{\Box\Obj\pvar{P}}[@]$.

        Feature for correct answer: your !!{structure} needs to be 
        reflexive (every world accesses itself) and have a failure of 
        transitivity. That is, three worlds such that the first 
        accesses the second, the second the third, but not the first
        the third. 

        \item A $\Log{D}$ !!{structure} that is not a $\Log{T}$
        !!{structure}.

        $\mModel{M}$ with $W=\{@,u\}$,
        $R=\{\tuple{@,@},\tuple{u,@}\}$, $V$ any valuation.
        $\mModel{M}$ is a $\Log{D}$ !!{structure} since $R$ is serial
        (for every $w$ in $W$, $w$ has access to some world). It is
        not a $\Log{T}$ !!{structure} because $R$ is not reflexive
        (because $u$ does not have access to itself).
        
        Feature for correct answer: your !!{structure} needs to be serial 
        (every world accesses some world) but not reflexive: some 
        world does not have access to itself. It's fine if \emph{some}
        but not all worlds access themselves, as in our answer above.

        \item A $\Log{S4}$ !!{structure} in which
        $\Diamond\Obj{\pvar{A}}\lif\Box\Diamond\Obj{\pvar{A}}$ is
        false.

        $\mModel{M}$ with $W=\{@,u\}$,
        $R=\{\tuple{@,@},\tuple{@,u},\tuple{u,u}\}$, 
        $V(\Obj{\pvar{A}})=\{@\}$.
        The model is reflexive. It's also transitive: for any worlds 
        $w,w',w''$ in the !!{structure}, if $Rww'$ and $Rw'w''$
        then $Rww''$. We could check all instances. Every instance 
        with $w=w'$ or $w'=w''$ is trivial:

        \smallskip
        \begin{tabular}{ccc}
            \hline
            $Rww'$ is & $Rw'w''$ is     & we have $Rww''$: \\
            \hline
            $R@@$ & $R@@$ & $R@@$ \\
            $R@@$ & $R@u$ & $R@u$ \\
            $R@u$  & $Ruu$ & $R@u$ \\
            \hline
        \end{tabular}
        \smallskip

        With only two worlds the only way transitivity would fail 
        would be to have $R@u$ and $Ru@$ but not $R@@$ or not $Ruu$.
        Since we don't have $Ru@$ this does not arise here; 
        the !!{structure} is transitive. 

        Since the !!{structure} is reflexive and transitive, it is a
        $\Log{S4}$ !!{structure}. Since $R@@$ and
        $\mSat{M}{\Obj{\pvar{P}}}[@]$,
        $\mSat{M}{\Diamond\Obj{\pvar{P}}}[@]$. Since $u$ only has 
        access to $u$ and $\mSat/{M}{\Obj{\pvar{P}}}[u]$,
        $\mSat/{M}{\Diamond\Obj{\pvar{P}}}[u]$. Since $R@u$ and 
        $\mSat/{M}{\Diamond\Obj{\pvar{P}}}[u]$, 
        $\mSat/{M}{\Box\Diamond\Obj{\pvar{P}}}[@]$.
        
        Feature for correct answer: a !!{structure} that is reflexive
        and transitive but where symmetry fails at $@$, that is, $@$
        has access to some world $u$ that doesn't have access to $@$,
        and $\Obj{\pvar{P}}$ is true at $@$ but not $u$ nor any world
        accessible from $u$. For transitivity, we must ensure that any
        sequence of access steps ($Rw_1w_2$, $Rw_2w_3$, \dots,
        $Rw_{n-1}w_n$) has a one-step shortcut ($Rw_1w_n$). This
        means, in particular, that $u$ isn't at the start of a chain
        of access steps that loops back to $@$.  

        \item A $\Log{S4}$ !!{structure} in which
        $\Diamond\Box\Obj{\pvar{A}}\lif\pvar{A}$ is false.

        Use the same frame as in the previous answer with
        $V(\Obj{\pvar{A}})=\{u\}$.

        Feature for correct answer: a !!{structure} that is reflexive
        and transitive but where symmetry fails at $@$, that is, $@$
        has access to some world $u$ that doesn't have access to $@$,
        and $\Obj{\pvar{P}}$ is false at $@$ but true at $u$ and any
        world accessible from $u$.

        \item A $\Log{B}$ !!{structure} in which
        $\Diamond\Obj{\pvar{A}}\lif\Box\Diamond\Obj{\pvar{A}}$ is
        false.

        $\mModel{M}$ with $W=\{@,u,v\}$,
        $R=\{\tuple{@,@},\tuple{u,u},\tuple{v,v},
        \tuple{@,u},\tuple{u,@},\tuple{@,v},\tuple{v,@}
        \}$, $V(\Obj\pvar{A})=\{u\}$.

        The !!{structure} is reflexive and symmetric. Reflexive:
        $\tuple{w,w}\in R$ for all $w\in W$. Symmetric: for any $w,w'$
        in $W$, if we have $\tuple{w,w'}\in R$ then we have
        $\tuple{w',w}\in R$ too (witness $@,v$ and $@,u$; the cases
        with $w=w'$ are trivial.) So it is a $\Log{B}$ !!{structure}.
        We have $\mSat{M}{\Diamond\Obj{\pvar{A}}}[@]$
        (since $R@u$ and $\mSat{M}{\Obj{\pvar{A}}}[u]$). But we also 
        have $\mSat/{M}{\Diamond\Obj{\pvar{A}}}[v]$ since $v$ only 
        has access to $@$ and $v$ and $@,v\notin V(\Obj{\pvar{A}})$;
        and since $R@v$ and $\mSat/{M}{\Diamond\Obj{\pvar{A}}}[v]$,
        $\mSat/{M}{\Box\Diamond\Obj{\pvar{A}}}[@]$.

        Feature for correct answer: a reflexive and symmetric
        !!{structure} where euclideanity fails at $@$, that is $@$ has
        access to two worlds $w_1$, $w_2$ that don't see each other;
        $\Obj{\pvar{A}}$ is true at one of these worlds, say $w_1$,
        but false at $@$, at $w_2$ and at any world the latter has
        access to. Note that since we need to ensure that the model is 
        symmetric we need $w_1$ and $w_2$ to both be distinct from $@$,
        for by symmetry any world that $@$ accesses must access $@$
        in return. 

        \item A $\Log{S5}$ !!{structure} in which
        $\Obj{\pvar{A}}\lif\Box\Obj{\pvar{A}}$ is false.

        $\mModel{M}$ with $W=\{@,u\}$,
        $R=\Setabs{\tuple{w,w}}{w\in W}$ (i.e. $R$ is universal, any 
        world has access to itself; you could write that $R=W\times W$
        too; see 'Cartesian product' in set theory), and 
        $V(\Obj\pvar{A})=\{@\}$.

        Feature for correct answer: $R$ must be an equivalence 
        relation. That means that $W$ must be divisible in a 
        \emph{partition} of 'cells' such that each world has access to all and 
        only the worlds that are in its cell. $\Obj\pvar{A}$ is 
        true at $@$ but false at some world in $@$'s cell. 

    \end{enumerate}
    
    \end{ans}
    
\end{prob}


\begin{prob}
    Say whether the following statement is correct. If so, explain why,
    if not, why not. 
    \begin{quote}
       If $!A$ is valid in $\Log{T}$, then $!A$ is valid in $\Log{S4}$. 
        (If $\Entails[\Log{T}]!A$, then $\Entails[\Log{S4}]!A$.)
    \end{quote}

    \begin{ans}
        Suppose $\Entails[\Log{T}]!A$: $!A$ is true in all pointed
        $\Log{T}$ !!{structure}s. $\Log{S4}$ !!{structure}s are $\Log{T}$
        !!{structure}s, because their accessibility relation is
        reflexive. Therefore $!A$ is true in all pointed $\Log{S4}$
        !!{structure}s: $\Entails[\Log{S4}]!A$.
    \end{ans}

\end{prob}


\end{document}

