% Part: normal-modal-logic
% Chapter: natural-deduction 
% Section: relate-to-axiomatic

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{nml}{nd}{rax}

\olsection{Relation with axiomatic systems}

!!^{derivation}s for modal logic are often given by axiomatic systems.
In those Normality is captured by an axiom schema and a rule:
\begin{itemize}
    \item \emph{Distribution} axiom schema. $\Box(!A\lif!B)\lif(\Box!A\lif\Box!B)$.
    \item \emph{Necessitation} rule. If $\Proves!A$, then $\Proves\Box!A$.
\end{itemize}
Given Normality and Modus Ponens ($!A\lif!B,!A\Entails !B$) all
instances of Distribution are logical truths. And given Normality again, 
considering the case where $\Gamma$ is empty, the Necessitation rule
is sound too. In axiomatic systems, propositional logic plus the
Distribution schema and the Necessitation rule turn out to be 
enough to capture all the consequences of Normality. It is easy to see
that the schema and rule are both !!{derivable} 
from our $Box$\Ax{K} rule. Necessitation is a special case of it 
(with no $\Box!A,\Box!B$ premises); Distribution instances are 
straightforwardly derived too.

Stronger systems given by characteristic axiom schemas. These are
obviously equivalent to our rules:
\begin{itemize}
    \tagitem{prvDiamond}
        {\Ax{D}. $\Box!A\lif\Diamond!A$}
        {\Ax{D} $\Box!A\lif\lnot\Box\lnot!A$} 
    \item $\Box$\Ax{T}. $\Box!A\lif!A$
    \iftag{prvDiamond}{\\ $\Diamond$\Ax{T}. $!A\lif\Diamond!A$}{}
    \item $\Box$\Ax{4}. $\Box!A\lif\Box\Box!A$
    \iftag{prvDiamond}{\\ $\Diamond$\Ax{4}. $\Diamond\Diamond!A\lif\Diamond!A$}{}
    \item $\Box$\Ax{B}. $!A\lif\Box\Diamond!A$
    \iftag{prvDiamond}{\\ $\Diamond$\Ax{B}. $\Diamond\Box!A\lif!A$}{}
    \item $\Box$\Ax{5}. $\Diamond!A\lif\Box\Diamond!A$
    \iftag{prvDiamond}{\\ $\Diamond$\Ax{5}. $\Diamond\Box!A\lif\Box!A$}{}
 
\end{itemize}

\end{document}