% Part: non-classical-logics Chapter: free-logics Section: introduction

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{ncl}{frl}{int}

\olsection{Introduction}

First-order logic has strong existence commitments. The following
are all valid in first-order logic (with identity for the latter two):

\begin{enumerate} 
	\item $\lexists x (\Atom{\Obj F}{x} \lor \lnot \Atom{\Obj F}{x})$ 
	\item $\lexists x (\eq[x][x])$ 
	\item $\lexists x (\eq[a][x])$ 
\end{enumerate}

The first two require that something exists. They are valid because
!!{structure}s of first-order logic have a non-empty domain. They are !!{derivable} from the fact that in first-order logic, we can !!{derive} $\lexists x !A(x)$ from $!A{a}$, together with further facts. For the first, the relevant fact is that substitutions of !!{derivable} formulas of propositional logic formulas are derivable in first-order logic too. Since $!A \lor \lnot !A$ is !!{derivable} in propositional logic, substituting $Atom{\Obj F}{a}$ for $!A$, we have $\Atom{\Obj F}{a} \lor \lnot \Atom{\Obj F}{a}$ which is !!{derivable} in first-order logic. For the second, the relevant fact is that $\eq[x][x]$ is !!{derivable} in first-order logic with identity. 

The third claim requires that every !!{constant} of the language denotes something. It is valid because in first-order !!{structure}s, the interpretations assigns an object of the !!{domain} to each !!{constant}. It is !!{derivable} from the fact that $\eq[a][a]$ is !!{derivable}, together with the fact that we can !!{derive} $\lexists x !A(x)$ from $!A{a}$.

These features of classical first-order logic are controversial because the following do not seem to \emph{logical} truths:

\begin{enumerate}
	\item There is something.
	\item Alice exists. 
\end{enumerate}

Logics have been devised to avoid these implications. They are:

\begin{enumerate}
	\item Inclusive logics, which allow the domain of quantification to be empty, and
	\item Free logics, which allow !!{constant}s that do not refer to any object in the domain of quantification.
\end{enumerate}

If our language includes !!{constant}s, then an inclusive logic must be free. So while inclusive logics need not be free, we will not discuss inclusive logics separately. 

Free logics can be inclusive or not. It is common to make free logics inclusive as well, so we'll focus on free logics that are also inclusive.

\end{document}
